---
title: "Lab 3 - Parallelizing k-means Stat 215A, Fall 2018"
author: "Todd Faulkenberry"
date: "20 October 2018"
header-includes:
   - \usepackage{float}
output: 
  pdf_document:
    number_sections: true
---

```{r, echo = FALSE, messages = FALSE}
# Loading necessary package
suppressMessages(library(tidyverse))
library(foreach)
library(parallel)
library(doParallel)
library(Rcpp)


```




```{r, echo = F}
# Loading in load.R and clean.R files
source('R/load.R')
source('R/clean.R')
source('R/functions.R')


# Loading and cleaning data
ling.binary <- loadRData()

```

## Introduction

In this project, we will implement an analyze an algorithim for computational stability that was pioneered by Ben Hur et. al @benhur. During our analysis, we will discuss the choice of similarity measure, compare C++ and R versions of the algorithim, and use the algorithm to see how many clusters it suggests. We will end the paper with a more general discussion of the algorithm and if it seems trustworthy after our analysis.

## Discussion of choice of similarity matrix

### Compare C++ and R versions of the similarity matrix




## Discussion of the algorithm

### Plot from part 3

### Discuss how many clusters to choose

### Discuss whether you trust the method or not


## Bibliography

---
references:
- id: benhur
  title: A stability based method for discovering structure in clustered data
  author:
  - family: Ben-Hur et. al.
    given: Asa
  URL: 'http://www.cs.colostate.edu/~asa/papers/psb02.pdf'
  publisher: Pacific Symposium on Biocomputing
  page: 6-17
  type: article-journal
  issued:
    year: 2002
---



